---
title: "Predictive Probability of Success (PoS)"
subtitle: "Virtual ICB case study: Using PFS to predict OS (log(HR), p<0.05)"
author: "Internal Methodology Share"
date: "2026-01-07"

format:
  beamer:
    aspectratio: 169
    fontsize: 10pt
    incremental: false
    include-in-header: |
      % Make code/output blocks smaller and allow line-wrapping in PDF slides
      \usepackage{fvextra}
      \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,breakanywhere,commandchars=\\\{\}}
      \AtBeginEnvironment{Shaded}{\footnotesize}
      \setlength{\emergencystretch}{3em}
  revealjs:
    theme: simple
    slide-number: true
    incremental: false
    center: false
    width: 1280
    height: 720
    html-math-method: mathjax
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    number-sections: true

execute:
  echo: false
  warning: false
  message: false
  cache: false
---
## Background: why PoS (not only power)

In drug development, a Go/No-Go decision often depends on the chance that a *future* confirmatory analysis will meet a pre-specified success criterion.

Classical power assumes the true treatment effect is fixed and known (or set to a design alternative). In practice, the true effect is uncertain and should be updated by accumulating evidence (historical trials, interim data, surrogate endpoints).

**Predictive Probability of Success (PoS)** answers:

> “Given what we know *today*, what is the probability the future study will be declared successful?”

---

## Core objects and notation

Let $Y^*$ denote the test statistic or effect estimate from the *future* (or not-yet-mature) analysis, and let $R$ be the **success region** defined by the protocol.

Examples of $R$:
- Frequentist: $p < 0.05$ (two-sided) and effect in the favorable direction  
- Bayesian: $\Pr(\theta < 0 \mid \text{data}) > 0.95$  
- Clinical: lower CI bound exceeds a clinically meaningful threshold

Let $\theta$ denote the **true treatment effect** (e.g., log hazard ratio).

---

## PoS definition: integral over uncertainty in the true effect

The building block is the conditional success probability:

$$
\Pr(Y^* \in R \mid \theta).
$$

Because $\theta$ is unknown, PoS averages the conditional success probability over the current posterior:

$$
\text{PoS}
= \int \Pr(Y^* \in R \mid \theta)\, p(\theta \mid Y_o)\, d\theta,
$$

where $Y_o$ denotes all evidence observed up to now (e.g., historical trials, interim information).

---

## Equivalent view: posterior predictive distribution

An equivalent expression uses the posterior predictive distribution of the future result:

$$
p(y^* \mid Y_o)
= \int p(y^* \mid \theta)\, p(\theta \mid Y_o)\, d\theta.
$$

Then PoS is simply:

$$
\text{PoS} = \Pr(Y^* \in R \mid Y_o)
= \int \mathbf{1}(y^* \in R)\, p(y^* \mid Y_o)\, dy^*.
$$

This view is often convenient for computation and interpretation.

---

## Success rule for log(HR) with two-sided $p<0.05$

Assume the future OS analysis estimates the log hazard ratio:

$$
Y^* \mid \theta \sim N(\theta, \sigma_*^2),
$$

and we target benefit with $HR < 1$ (i.e., $\theta < 0$).

For a two-sided test at $\alpha = 0.05$, the success region can be written as:

$$
R = \{\, Y^* < c \,\}, \quad c = -z_{1-\alpha/2}\, \sigma_* = -1.96\,\sigma_*.
$$

Therefore:

$$
\text{PoS} = \Pr(Y^* < c \mid Y_o).
$$

---

## Linking the ICB surrogate scenario to PoS

In the ICB setting, PFS typically matures earlier than OS.

We use historical trials reporting both endpoints to learn the joint relationship:

$$
(\theta_{OS}, \theta_{PFS})^\top \sim \mathrm{MVN}(\mu,\Sigma_{\text{between}}),
$$

and then update OS using the observed current-trial PFS result $y_c$.

This yields a predictive distribution for the final OS estimate:

$$
Y^*_{OS} \mid y_c \sim N\!\left(m_{\text{post}},\, v_{\text{post}} + \sigma_*^2\right),
$$

and PoS follows from the OS success rule:

$$
\text{PoS} = \Pr(Y^*_{OS} < -1.96\,\sigma_* \mid y_c).
$$


## What we’re trying to answer

**Scenario (ICB trial):**

- Primary endpoint: **PFS** (read out earlier)
- Key secondary endpoint: **OS** (matures later)
- Effect measure: **log(HR)** for both endpoints

**Decision question:**

> Given the current trial’s observed PFS result, what is the probability that the final OS analysis will be statistically significant (two-sided p < 0.05) in the beneficial direction?

::: notes
This is the core “PoS” story: translate earlier evidence (PFS) into an actionable probability about OS success, while carrying uncertainty.
:::

---

## PoS in one line

Let:

- $\theta$ = the (unknown) *true* treatment effect (here: $\theta = \log(HR)$)
- $Y^*$ = the future study’s estimator (e.g., estimated log(HR) at final OS analysis)
- $R$ = the **success region** (defined by your decision rule)

Then predictive probability of success is:

$$
\text{PoS} \,=\, \int \Pr\big(Y^*\in R\mid \theta\big)\;\pi(\theta\mid \text{current evidence})\, d\theta
$$

**Interpretation:** average the conditional “power” over uncertainty in the true effect.

---

## Our success rule (log(HR), two-sided p < 0.05)

For OS final analysis, assume:

$$
Y^*_{OS} \mid \theta^*_{OS} \sim \mathcal N(\theta^*_{OS},\; se_{*}^2)
$$

Two-sided $\alpha=0.05$, beneficial direction $HR<1$ (so $\log(HR)<0$):

$$
R = \{Y^*_{OS} < c\},\quad c = -z_{0.975}\, se_{*} \approx -1.96\, se_{*}
$$

So:

$$
\text{PoS} = \Pr\left(Y^*_{OS} < c\mid \text{current evidence}\right)
$$

::: notes
This is just the usual z-test rejection region, written on the log(HR) scale.
:::

---

## Where “current evidence” comes from

We will combine two sources:

1) **Historical trials** (e.g., 27 ICB studies) that report both endpoints
   - $y_{i,OS}=\log(HR_{OS,i})$, $se_{i,OS}$
   - $y_{i,PFS}=\log(HR_{PFS,i})$, $se_{i,PFS}$

2) **Current trial’s PFS** (available now)
   - $y_c = \log(HR_{PFS})$, $se_c$

Goal: obtain $\theta^*_{OS}\mid y_c$ (predictive distribution for the current trial’s *true* OS effect), then compute PoS.

---

## Model (concept picture)

Two layers:

- **Between-study layer (true effects differ across trials):**

$$
(\theta_{i,OS},\theta_{i,PFS})^\top \sim \text{MVN}(\mu,\,\Sigma_{between})
$$

- **Within-study layer (estimation error):**

$$
(y_{i,OS},y_{i,PFS})^\top \sim \text{MVN}((\theta_{i,OS},\theta_{i,PFS})^\top,\,\Sigma_{within,i})
$$

Key parameters:

- $\mu_{OS},\mu_{PFS}$: average (logHR) effects across trials
- $\tau_{OS},\tau_{PFS}$: between-study SDs (heterogeneity)
- $\rho$: correlation between true OS and true PFS effects

::: notes
The whole approach rests on “exchangeability”: the historical trials and the current trial are assumed to come from the same super-population (up to heterogeneity).
:::

---

## Conditioning: OS effect given observed PFS

Treat the current trial’s true effects $(\theta^*_{OS},\theta^*_{PFS})$ as a draw from the same between-study distribution.

Observation model for current PFS:

$$
y_c = \theta^*_{PFS} + e,\quad e\sim \mathcal N(0,se_c^2)
$$

If $(\theta^*_{OS},\theta^*_{PFS})\sim \text{MVN}(\mu,\Sigma_{between})$, then:

$$
\theta^*_{OS}\mid y_c \sim \mathcal N(m_{post}, v_{post})
$$

where

$$
m_{post} = \mu_{OS} + \frac{\text{Cov}(OS,PFS)}{\text{Var}(PFS)+se_c^2}\,(y_c-\mu_{PFS})
$$
$$
v_{post} = \text{Var}(OS) - \frac{\text{Cov}(OS,PFS)^2}{\text{Var}(PFS)+se_c^2}
$$

---

## PoS in closed form

Final OS estimator:

$$
Y^*_{OS}\mid y_c \sim \mathcal N\big(m_{post},\; v_{post}+se_{*}^2\big)
$$

Therefore:

$$
\text{PoS} = \Pr(Y^*_{OS}<c\mid y_c)
= \Phi\!\left(\frac{c-m_{post}}{\sqrt{v_{post}+se_{*}^2}}\right)
$$

**Knobs that matter:**

- $\rho$: stronger PFS–OS association → PFS updates OS more
- $\tau$’s: more heterogeneity → less certainty → PoS often down
- $d_{events}$: more OS events → smaller $se_{*}$ → PoS up

---

## Reproducible demo: simulate 27 historical ICB trials

We’ll create a virtual dataset with:

- K = 27 historical trials
- True bivariate relationship between log(HR_OS) and log(HR_PFS)
- Realistic ranges of standard errors

```{r}
#| label: setup
#| include: false

set.seed(20260106)

# Wrap console output to avoid going off-slide in PDF
options(width = 70)

make_pd <- function(S, eps = 1e-6) {
  dn <- dimnames(S)
  e <- eigen(S, symmetric = TRUE)
  vals <- pmax(e$values, eps)
  S_pd <- e$vectors %*% diag(vals) %*% t(e$vectors)
  dimnames(S_pd) <- dn
  S_pd
}

library(MASS)

K <- 27

# "True" hyper-parameters for the virtual historical world
mu_true <- c(OS = -0.20, PFS = -0.30)        # mean log(HR)
tau_true <- c(OS = 0.15, PFS = 0.12)         # between-study SDs
rho_true <- 0.75

Sigma_between_true <- matrix(
  c(tau_true["OS"]^2,
    rho_true*tau_true["OS"]*tau_true["PFS"],
    rho_true*tau_true["OS"]*tau_true["PFS"],
    tau_true["PFS"]^2),
  nrow = 2, byrow = TRUE,
  dimnames = list(c("OS","PFS"), c("OS","PFS"))
)

# Standard errors per trial/endpoints (roughly realistic)
se_os  <- runif(K, 0.10, 0.18)
se_pfs <- runif(K, 0.08, 0.15)

# Within-study correlation between (y_OS, y_PFS) often unknown; set 0 for base demo
r_within <- 0
Sigma_within_list <- lapply(1:K, function(i) {
  matrix(c(se_os[i]^2, r_within*se_os[i]*se_pfs[i],
           r_within*se_os[i]*se_pfs[i], se_pfs[i]^2),
         nrow = 2, byrow = TRUE)
})

# Simulate true effects and observed estimates
theta <- mvrnorm(n = K, mu = mu_true, Sigma = Sigma_between_true)
colnames(theta) <- c("OS","PFS")

y <- matrix(NA_real_, nrow = K, ncol = 2, dimnames = list(NULL, c("OS","PFS")))
for (i in 1:K) {
  e_i <- mvrnorm(n = 1, mu = c(0,0), Sigma = Sigma_within_list[[i]])
  y[i,] <- theta[i,] + e_i
}

hist_dat <- data.frame(
  trial = 1:K,
  y_os = y[,"OS"],  se_os = se_os,
  y_pfs = y[,"PFS"], se_pfs = se_pfs
)

head(hist_dat, 5)
```

---

## Estimate the between-study relationship (demo-friendly moment approach)

For a quick internal demo, we’ll use **error-corrected moments**:

- observed sample covariance: $S_{obs}=\text{Cov}(y)$
- average within-study covariance: $\bar W$
- estimate: $\hat\Sigma_{between} \approx S_{obs}-\bar W$

```{r}
#| label: estimate_between

S_obs <- cov(y)
W_bar <- Reduce("+", Sigma_within_list) / K

Sigma_between_hat <- make_pd(S_obs - W_bar)
mu_hat <- colMeans(y)

tau_hat <- sqrt(diag(Sigma_between_hat))
rho_hat <- Sigma_between_hat["OS","PFS"] / (tau_hat[1]*tau_hat[2])

list(
  mu_hat = mu_hat,
  Sigma_between_hat = Sigma_between_hat,
  tau_hat = tau_hat,
  rho_hat = rho_hat
)
```

::: notes
For production work you’d typically fit a full bivariate random-effects model (REML or Bayesian). This moment approach is intentionally simple for a clean demo.
:::

---

## Current trial: plug in observed PFS and compute PoS

Assume the current trial has interim/final PFS result:

- $y_c = \log(HR_{PFS})$
- $se_c$ (its standard error)

For OS final analysis, assume planned OS event count $d$.
Under 1:1 allocation, Schoenfeld gives a useful approximation:

$$
se_{*} \approx \frac{2}{\sqrt{d}}
$$

```{r}
#| label: pos_calc

# Current PFS (modify to explore scenarios)
y_c  <- -0.35   # log(HR_PFS)  ~ HR 0.70
se_c <-  0.09

# Planned OS final analysis events
# d_events = number of OS events (e.g., deaths) at the final analysis
# For 1:1 allocation, se* ≈ 2/sqrt(d)
d_events <- 240
se_star <- 2 / sqrt(d_events)

# Success threshold for two-sided alpha=0.05 and beneficial direction
c_thr <- -qnorm(0.975) * se_star

# Conditional distribution theta*_OS | y_c
var_pfs <- Sigma_between_hat["PFS","PFS"]
var_os  <- Sigma_between_hat["OS","OS"]
cov_op  <- Sigma_between_hat["OS","PFS"]

m_post <- mu_hat["OS"] + cov_op/(var_pfs + se_c^2) * (y_c - mu_hat["PFS"])
v_post <- var_os - cov_op^2/(var_pfs + se_c^2)

sd_pred <- sqrt(v_post + se_star^2)
PoS <- pnorm(c_thr, mean = m_post, sd = sd_pred)

c(
  y_c = y_c,
  HR_PFS = exp(y_c),
  m_post_OS = m_post,
  HR_OS_pred_mean = exp(m_post),
  sd_thetaOS = sqrt(v_post),
  d_events = d_events,
  se_star = se_star,
  threshold_c = c_thr,
  PoS = PoS
)
```

---

## Visual 1: historical relationship (virtual 27 trials)

```{r}
#| label: plot_scatter
#| fig-height: 4.6
#| fig-width: 8.5

plot(hist_dat$y_pfs, hist_dat$y_os,
     xlab = "Historical log(HR_PFS)",
     ylab = "Historical log(HR_OS)",
     main = "Virtual 27 ICB trials: log(HR_PFS) vs log(HR_OS)",
     pch = 16)

abline(h = 0, v = 0, lty = 2)
fit_lm <- lm(y_os ~ y_pfs, data = hist_dat)
abline(fit_lm, lwd = 2)

points(y_c, m_post, pch = 17, cex = 1.3)
text(y_c, m_post, labels = "  current trial (PFS → OS mean)", pos = 4)

legend("topright",
       legend = c("Historical trials", "OLS line (visual only)", "Current trial (mapped mean)"),
       pch = c(16, NA, 17), lty = c(NA, 1, NA), bty = "n")
```

---

## Visual 2: PoS curve (how PFS maps to OS success probability)

```{r}
#| label: plot_pos_curve
#| fig-height: 4.6
#| fig-width: 8.5

grid_yc <- seq(-0.70, 0.05, length.out = 220)
PoS_grid <- sapply(grid_yc, function(yc) {
  m <- mu_hat["OS"] + cov_op/(var_pfs + se_c^2) * (yc - mu_hat["PFS"])
  v <- var_os - cov_op^2/(var_pfs + se_c^2)
  sd <- sqrt(v + se_star^2)
  pnorm(c_thr, mean = m, sd = sd)
})

plot(grid_yc, PoS_grid, type = "l",
     xlab = "Current observed log(HR_PFS)  (y_c)",
     ylab = "PoS for OS (two-sided p<0.05, HR<1)",
     main = "PoS curve: PFS result → OS success probability")

abline(v = y_c, lty = 2)
points(y_c, PoS, pch = 16)
text(y_c, PoS, labels = sprintf("  y_c=%.2f, PoS=%.2f", y_c, PoS), pos = 4)
```

::: notes
This plot is what decision-makers usually want: a quantitative map from current PFS readout to probability of final OS success.
:::

---

## Sensitivity demo: vary PFS–OS correlation ($\rho$)

```{r}
#| label: sens_rho
#| fig-height: 4.6
#| fig-width: 8.5

rho_grid <- c(0.50, 0.75, 0.90)

plot(NULL, xlim = range(grid_yc), ylim = c(0,1),
     xlab = "Current observed log(HR_PFS)  (y_c)",
     ylab = "PoS for OS",
     main = "Sensitivity: PoS curve under different rho")

for (rho_j in rho_grid) {
  Sigma_j <- Sigma_between_hat
  Sigma_j["OS","PFS"] <- rho_j * sqrt(Sigma_j["OS","OS"]*Sigma_j["PFS","PFS"])
  Sigma_j["PFS","OS"] <- Sigma_j["OS","PFS"]
  Sigma_j <- make_pd(Sigma_j)

  var_pfs_j <- Sigma_j["PFS","PFS"]
  var_os_j  <- Sigma_j["OS","OS"]
  cov_op_j  <- Sigma_j["OS","PFS"]

  PoS_j <- sapply(grid_yc, function(yc) {
    m <- mu_hat["OS"] + cov_op_j/(var_pfs_j + se_c^2) * (yc - mu_hat["PFS"])
    v <- var_os_j - cov_op_j^2/(var_pfs_j + se_c^2)
    sd <- sqrt(v + se_star^2)
    pnorm(c_thr, mean = m, sd = sd)
  })
  lines(grid_yc, PoS_j, lwd = 2)
}

abline(v = y_c, lty = 2)
legend("topleft", legend = paste0("rho=", rho_grid), lwd = 2, bty = "n")
```

---

## Sensitivity demo: OS information (event count) matters

If your event count changes, so does $se_{*}$ and the success threshold $c$.

```{r}
#| label: sens_events

event_grid <- c(150, 200, 240, 300, 400)
res <- data.frame(d_events = event_grid, se_star = 2/sqrt(event_grid))
res$threshold_c <- -qnorm(0.975) * res$se_star

# Recompute PoS at the fixed current y_c for each event count
res$PoS <- sapply(event_grid, function(d) {
  se_s <- 2/sqrt(d)
  c_s  <- -qnorm(0.975) * se_s
  sd_p <- sqrt(v_post + se_s^2)
  pnorm(c_s, mean = m_post, sd = sd_p)
})

res
```

---

## Takeaways (what to emphasize internally)

- PoS answers a **decision** question: *probability of meeting a success rule*, not “best guess HR”.
- In this ICB setup:
  - Historical trials estimate a **between-study** OS–PFS association ($\rho$, heterogeneity)
  - Current PFS readout updates the predicted **true OS effect** via conditional normal formulas
  - Final OS success probability is a **one-line Normal CDF**
- Sensitivity analysis is not optional:
  - $\rho$ (association strength), heterogeneity ($\tau$), and OS information (event count) are the big levers.

---

## Next steps (toward production)

If this moves beyond a virtual demo:

- Fit a full bivariate random-effects model (REML/Bayesian)
- Treat within-study correlation $r$ as unknown → range or prior
- Add design features that affect OS–PFS link (crossover, line of therapy, combo vs mono)
- Use the same framework for interim decisions: re-run PoS as PFS/OS mature

::: notes
For an internal share, you can end with a clear checklist: what you need to collect from historical trials and what assumptions drive sensitivity.
:::

---

## How to run this deck

1) Save this file as `PoS_virtual_ICB_demo.qmd`
2) In RStudio / VS Code terminal:

```bash
quarto render PoS_virtual_ICB_demo.qmd
```

It will produce an HTML slide deck (RevealJS) with all plots and numbers.